---
title: "NRES 776 Lab 9 - GLM with single predictor"
author: "Your name here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Poisson regression - horseshoe crab

This problem refers to data from a study of nesting horseshoe crabs (J. Brockmann, Ethology 1996); see also Agresti (1996) Sec. 4.3 and Agresti (2002) Sec. 4.3. Each female horseshoe crab in the study had a male crab attached to her in her nest. The study investigated factors that affect whether the female crab had any other males, called satellites, residing near her. Explanatory variables that are thought to affect this included the female crab's color (C), spine condition (S), weight (Wt), and carapace width (W). The response outcome for each female crab is her number of satellites (Sa). There are 173 females in this study.

In this practice, we are going to explore the relationship between number of satellites (Sa) vs color (C) using Poisson regression.

## Load required libraries and data set

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(lmtest)

data_crab <- read_table(here("crab_dataset.txt")) 

data_crab_clean <- data_crab %>%
  mutate(C = factor(C, labels = c("Light", "Median", "Dark", "Very dark")))
  
```

```{r}
data_crab_clean
```

## Exploratory visualization - Sa vs C

Seems like the number of number of satellites (Sa) decreases as the color (C) gets darker. Interesting observation! We can use statistics to validate this observation.

```{r}
ggplot(aes(x = C, y = Sa), data = data_crab_clean) +
  geom_boxplot()
```

## Fit a Poisson regression

This is the model equation that we are going to fit. Explain why there are three parameters **(XYZ pt)**

$$
log(\mu_i) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 x_{3i}
$$

*Write your answer here*

Fit a Poisson using `glm()` function. Specify the arguments: `formula`, `data`, and `family`. We will use the default link function, which is a log link. **(XYZ pt)**

```{r}
### Write your answer here

# glm_pois_1 <- glm(___)
```

Then, use the `summary()` function to print the model output. **(XYZ pt)**

```{r}
### Write your answer here

# ___
```

The model output indicated the estimated values for the parameters, and model statistics. Report these values. (Hint: use `exp()` function and `coef()` function to get exponential value easier) **(XYZ pt)**

| parameter/statistics | estimated value          | exponential of estimated value |
|--------------------|------------------------|-----------------------------|
| $\beta_0$            | *Write your answer here* | *Write your answer here*       |
| $\beta_1$            | *Write your answer here* | *Write your answer here*       |
| $\beta_2$            | *Write your answer here* | *Write your answer here*       |
| $\beta_3$            | *Write your answer here* | *Write your answer here*       |
| dispersion parameter | *Write your answer here* | NA                             |
| AIC                  | *Write your answer here* | NA                             |

## Model goodness of fit

Is the model including colour as predictor variable better than a null model? If the model performs significantly better than a null model, then we keep the predictor variable; if not, then we say that the colour is not significantly associated with the number of satellite.

Hypothesis of the likelihood ratio test (in words) **(XYZ pt)**

$H_0$: *Write your answer here*

$H_a$: *Write your answer here*

First, we need to fit a null model **(XYZ pt)**

```{r}
### Write your answer here

# glm_pois_null <- glm(___)

```

Then, run a likelihood ratio test **(XYZ pt)**

```{r}
### Write your answer here

# ___(glm_pois_null, glm_pois_1)
```

What is your decision based on this likelihood ratio test? (Hint: provide the p-value, the $\alpha$ threshold you are comparing with, and your conclusion) **(XYZ pt)**

*Write your answer here*

## Model result interpretation

Calculate the predicted mean number of satellites for horsecrab with Light colour. Show the calculation process. **(XYZ pt)**

*Write your answer here*

Calculate the predicted mean number of satellites for horsecrab with Dark colour. Show the calculation process. **(XYZ pt)**

*Write your answer here*

Is your modelling result showing the same pattern as you observed in the boxplot? Is colour of female horseshoe crab significantly influencing the number of satellite? Explain why or why not, use the model output value as needed. **(XYZ pt)**

*Write your answer here*

## Binomial regression - titanic 

The Titanic data set contains information about 887 of the 2229 souls aboard the ill-fated ocean liner Titanic. Each passenger\'s record contains the following attributes:

-   Name

-   Passenger class (1/2/3)

-   Sex

-   Age

-   Whether the passenger was accompanied by siblings, parents or children

-   The Fare they paid, and most importantly,

-   **Whether they survived (1=Survived, 0=Died)**

In this practice, we want to know whether the survival probability (Survival) is influence by the passenger class (Pclass).

## Load required libraries and data set

```{r, warning=FALSE, message=FALSE}
data_titanic <- read_csv(here("titanic_dataset.csv"))   

data_titanic_clean <- data_titanic %>%
  mutate(Pclass = factor(Pclass, labels = c("First", "Second", "Third")))
```

```{r}
data_titanic_clean
```

## Exploratory visualization - Survival vs Pclass

Seems like the number of number of satellites (Sa) decreases as the color (C) gets darker. Interesting observation! We can use statistics to validate this observation.

```{r}
ggplot(aes(x = Survival, y = Pclass), 
       data = data_crab_clean) +   
  geom_boxplot()
```

## Fit a Poisson regression

This is the model equation that we are going to fit. Explain why there are three parameters **(XYZ pt)**

$$ log(\mu_i) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 x_{3i} $$

*Write your answer here*

Fit a Poisson using `glm()` function. Specify the arguments: `formula`, `data`, and `family`. We will use the default link function, which is a log link. **(XYZ pt)**

```{r} ### Write your answer here  # glm_pois_1 <- glm(___)}
```

Then, use the `summary()` function to print the model output. **(XYZ pt)**

```{r} ### Write your answer here  # ___}
```

The model output indicated the estimated values for the parameters, and model statistics. Report these values. (Hint: use `exp()` function and `coef()` function to get exponential value easier) **(XYZ pt)**

| parameter/statistics | estimated value          | exponential of estimated value |
|--------------------|------------------------|-----------------------------|
| $\beta_0$            | *Write your answer here* | *Write your answer here*       |
| $\beta_1$            | *Write your answer here* | *Write your answer here*       |
| $\beta_2$            | *Write your answer here* | *Write your answer here*       |
| $\beta_3$            | *Write your answer here* | *Write your answer here*       |
| dispersion parameter | *Write your answer here* | NA                             |
| AIC                  | *Write your answer here* | NA                             |

## Model goodness of fit

Is the model including colour as predictor variable better than a null model? If the model performs significantly better than a null model, then we keep the predictor variable; if not, then we say that the colour is not significantly associated with the number of satellite.

Hypothesis of the likelihood ratio test (in words) **(XYZ pt)**

$H_0$: *Write your answer here*

$H_a$: *Write your answer here*

First, we need to fit a null model **(XYZ pt)**

```{r} ### Write your answer here  # glm_pois_null <- glm(___)}
```

Then, run a likelihood ratio test **(XYZ pt)**

```{r} ### Write your answer here  # ___(glm_pois_null, glm_pois_1)}
```

What is your decision based on this likelihood ratio test? (Hint: provide the p-value, the $\alpha$ threshold you are comparing with, and your conclusion) **(XYZ pt)**

*Write your answer here*

## Model result interpretation

Calculate the predicted mean number of satellites for horsecrab with Light colour. Show the calculation process. **(XYZ pt)**

*Write your answer here*

Calculate the predicted mean number of satellites for horsecrab with Dark colour. Show the calculation process. **(XYZ pt)**

*Write your answer here*

Is your modelling result showing the same pattern as you observed in the boxplot? Is colour of female horseshoe crab significantly influencing the number of satellite? Explain why or why not, use the model output value as needed. **(XYZ pt)**

*Write your answer here*
