---
title: "NRES 776 Lab 9"
subtitle: "GLM with single predictor"
author: Sunny Tseng
format:
  revealjs: 
    slide-number: true
    preview-links: auto
    #css: styles.css
    #footer: <https://quarto.org>
    theme: [default, styles.scss]
    #smaller: true
    #scrollable: true
    incremental: true
    embed-resources: true
    width: 1200
    fontsize: 1.7em
editor: visual
---

```{r setup}
library(tidyverse)
library(here)
library(lmtest)

binomial_data <- read_csv(here("data", "isolation.csv"))

binomial_data <- binomial_data %>% 
  mutate(isolation = cut(isolation, breaks = c(2, 5, 7, 10), 
                         labels = c("low", "median", "high"))) %>%
  mutate(incidence = as_factor(incidence)) 

```

## Our schedule today

-   Discussion paper presentation (50 min)

-   Binomial regression with single continuous $x$ (15 min)

-   Lab intro - Poisson regression with single predictor variable (10 min)

-   Lab intro - Binomial regression with single predictor variable (10 min)

-   Practice (60 min)

# Discussion paper presentation

-   Olivia Wilson

-   Tina Watters

-   Brett Gandy

# Binomial regression

## Research question (1 continuous x)

A survey was done on 50 islands for the incidence of a bird species Grasshopper Warbler. Researchers want to know whether the incidence is related to the area and/or the isolation level of the islands.

::: columns
::: {.column width="50%"}
### Use long format as input

```{r}
data_long <- binomial_data
data_long
```
:::

::: {.column width="50%"}
### Take a look at the relationship

```{r}
data_wide <- binomial_data %>%   
  group_by(isolation) %>%
  summarize(presence = sum(incidence == 1),  
            absence = sum(incidence == 0), 
            total = n(),                   
            proportion = presence/total,  
            mean_area = mean(area)) %>%     
  select(proportion, total, mean_area)   
data_wide
```
:::
:::

## Model formulation

$$ logit(p_i) = \beta_0 + \beta_1 x_{1i} $$

```{r, echo = TRUE}
incidence_binom <- glm(formula = incidence ~ area, 
                       data = data_long,  
                       family = "binomial")  
incidence_binom %>% summary
```

## Coef. interpretation

-   For island with area as 0 (baseline):

$$ logit(p_i) = log(\frac{p_i}{1-p_i})= \beta_0 $$

$$ \frac{p_i}{1-p_i} = Odd(p_i) = exp(\beta_0) = exp(-2.15) = 0.11 $$

-   The odd for the intercept is not often interpreted by itself.

## Coef. interpretation (con'd)

-   For island with non-0 area:

$$ logit(p_i) = \beta_0 + \beta_1 x_{1i}$$

$$ logit(p_i') = \beta_0 + \beta_1 (x_{1i} + 1)$$

$$ logit(p_i') - logit(p_i) = \beta_1 = log(\frac{Odd(p_i')}{Odd(p_i)}) $$

$$ \frac{Odd(p_i')}{Odd(p_i)} = exp(\beta_1) = exp(0.62) = 1.87  $$

-   For one unit increase in `area`, the odds of the bird species being present increase by a factor of 1.87.

## Output interpretation

### Odd ratio (OR)

-   $OR = 1$, no difference between groups
-   $OR < 1$, treatment decreases odds
-   $OR > 1$, treatment increases odds

### R output

-   **Intercept (-2.15)**: The odd of the bird being present on 0 area island is $exp(-2.15)$

-   **area (0.62)**: For one unit increase in `area`, the odds of the bird species being present increase by a factor of $exp(0.62)$

-   **Dispersion parameter (1)**: wonderful. Need to consider other methods if dispersion larger than 1 (over-dispersion) or smaller than 1 (under-dispersion)

-   **AIC (54.172)**: Can be used to compare the goodness of fit between models

## Model goodness of fit: Likelihood ratio test

```{r, echo = TRUE}
incidence_binom_null <- glm(formula = incidence ~ 1,
                            data = data_long,  
                            family = "binomial")
```

$H_0$: The model performance is the same as a null model (making predictions by chance)

$H_1$: The model performance is significantly different comparing to a null model

. . .

-   Use `lrtest()` function in the `lmtest` package

```{r, echo = TRUE}
lrtest(incidence_binom, incidence_binom_null)
```

## Model prediction

-   Use `predict()` and specify `type = "response"` to get back transformed $y_i$

```{r, echo = TRUE}
incidence_p <- data_long %>% 
  mutate(incidence_p = predict(incidence_binom, type = "response")) 
incidence_p
```

## Model visualization

```{r, echo = TRUE}
data_long %>%   
  mutate(incidence = incidence %>% as.numeric() - 1) %>%  
  ggplot(aes(x = area, y = incidence), data = .) +   
  geom_jitter(width = 0, height = 0.05) +  
  geom_smooth(method =  "glm",                
              method.args = list(family = "binomial"), 
              se = FALSE) +   
  labs(y = "Probability of incidence", x = "Area of an island")   
```

## Model visualization

```{r, echo = TRUE}
ggplot(aes(x = isolation, y = incidence_p),
       data = incidence_p) +   
  geom_boxplot() +   
  labs(y = "Probability of incidence", x = "Isolation level of island")
```

## Lab 9 - GLM with single predictor

## Get the lab assignment file

1.  Download the file on Moodle `Lab 9 - assignment file`. You should get a `.Rmd` file after downloading
2.  Open RStudio
3.  Install and load `rmarkdown` package

```{r, echo = TRUE, eval = FALSE}
install.packages("rmarkdown")
library(rmarkdown)
```

4.  Open the `.Rmd` file in Rstudio
5.  Click `knit` and you should get a `html` output
6.  You are all set!

## Submission

### Due next Wednesday 23:59 pm, Moodle

### Evaluation

-   File name and type (NRES776\_*firstname*\_*lastname*\_lab_9.**html**) (20%)

-   Accuracy of code (60%)

-   Readability, efficiency, and annotation of code (20%)
