---
title: "NRES 776 Syllabus"
author: "Lisa Koetke & Sunny Tseng"
date: "Fall 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(here)
```

### Advanced statistical analyses for natural resource sciences

- **Lecture**: Tuesday and Thursday 12:30 - 13:20 pm
- **Lab**: Thursday 8:00 - 10:50 am
- **Location**: In person: 5-177 Lecture & 8-362 Lab; Virtual: [zoom link](https://unbc.zoom.us/j/64078816232?pwd=TytUa2pnR1AyTmFYd2p5TXcrajhuZz09)
- **Course website**: UNBC Moodle (everything is there!)


This course provides advanced, practical training in the analysis of quantitative data. The course focuses primarily on advanced univariate statistics that can be applied to both experimental and observational data. This focus includes a full exploration of generalized linear models, including linear, logistic, and mixed models. Students also receive instruction in the principles of experimental design, data management, and the review and reporting of statistical results. Labs focus on using R statistical software to manipulate, display, and analyze data.


### Instructors


```{r echo = FALSE, results = "asis", message=FALSE}
instructor <- tibble(name = c("Lisa Koetke", "Sunny Tseng"),
                     email = c("lisa.koetke@unbc.ca", "stseng@unbc.ca"),
                     office = c("10-2086", "https://unbc.zoom.us/j/6088522970?pwd=cjJhOVRqMmk0bkJ1N2tJajJrTWNYdz09"),
                     `office hour` = c("Tue. 13:30 - 14:30", "Tue. 13:30 - 14:30"))
kable(instructor)

```




### Schedules

#### Lecture schedule

Lecture content will introduce the principles of experimental design and discuss a variety of statistical methods used to
analyze experimental and observational ecological data.

```{r echo = FALSE, results = "asis", message=FALSE}
lecture_schedule <- read_csv(here("docs", "lecture_schedule.csv"))
kable(lecture_schedule)

```

#### Lab schedule

The labs will include a tutorial demonstrating code and techniques, practice exercises, and a discussion of the weeks’ readings. Labs will emphasize hands-on statistical analysis and interpretation of results from quantitative observations and manipulative experiments. Analysis will be done using R Statistical Software.

```{r echo = FALSE, results = "asis", message=FALSE}
lab_schedule <- read_csv(here("docs","lab_schedule.csv"))
kable(lab_schedule)

```


### Assessments of learning

#### Grade distribution

```{r echo = FALSE, results = "asis", message=FALSE}
grades <- tibble(Letter = "Percent",
                 "A+" = "90.0 - 100.0",
                 "A" = "85.0 - 89.9",
                 "A-" = "80.0 - 84.9",
                 "B+" = "77.0 - 79.9",
                 "B" = "73.0 - 76.9",
                 "B-" = "70.0 - 72.9",
                 "C+" = "67.0 - 69.9",
                 "C" = "63.0 - 66.9",
                 "C-" = "60.0 - 62.9",
                 "D+" = "57.0 - 59.9",
                 "D" = "53.0 - 56.9",
                 "D-" = "50.0 - 52.9",
                 "F" = "<50.0")
kable(grades)

```


#### Lab assignments (50%)
- Due: every Thu. midnight
- Submission: a R file named as "NRES776_*lastname*_lab_*X*.R". Submit to course website.
- Description: For these assignments, you will analyze
data sets using a variety of techniques learned in class. A demonstration of lab assignments will be included in the beginning of each lab. Students are encouraged to finish the lab during the course lab time.

#### Project 1 (15%)
- Due: Oct.19 at midnight
- Submission: oral presentation, a PDF file named as "NRES776_*lastname*_project_1.pdf" submitted to course website.
- Description: Working with the instructor, students will develop an analysis strategy to examine data relevant to their research area. You will develop the project starting with a research question, data visualization, and list hypotheses. Making it as a research proposal that you are going to apply funding with. There will be a 5 mins presentation component as well as a 2-pages report.


#### Project 2 (20%)
- Due: Dec. 7 at midnight
- Submission: oral presentation during lecture times Nov. 28 to Dec. 5, a PDF file named as "NRES776_*lastname*_project_2.pdf" submitted to course website.
- Description: Your second project will build off of Project 1. Using the same dataset, you will answer your research questions and test your hypotheses using methods learned in the second portion of the semester. This project will include a 5-minute presentation and a written report of your methods and results.


#### Discussion paper presentation (10%)
- Due: one of the lab session
- Submission: oral presentation
- Description: You will have 5 mins to present a discussion paper during the lab period and lead a 20 mins discussion based on the papers. You can sign up for the paper you would like to present. All students are expected to read the discussion paper before each lab and participate discussion. Guidelines on leading a discussion will be provided.

#### General participation (5%)
- Description: You will be evaluated by your involvement in course discussion, and providing feedback to other students' presentation. 



### Resources

There is no required textbooks for this course. If you would like to have stats books on hand, here are a few useful ones. There are also many free, open source resources available for learning statistics online. 

#### Books

- Whitlock, MC and Schluter, D. 2015. The Analysis of Biological Data. 3rd Edition. Roberts and Company
Publishers.

- Gotelli, N.J., and A.M. Ellison. 2013. A Primer of Ecological Statistics (2nd Edition). Sinauer Associates
Inc., Sunderland, MA.

- Zuur, AF, Ieno, EN and Smith, GM. 2007. Analysing Ecological Data. Springer. eBook available through
UNBC library: http://prxy.lib.unbc.ca/login?url=https://search.ebscohost.com/login.aspx?direct=true&db=nlebk&AN= 254638&site=ehost-live&scope=site

#### Online resources

- R for Data Science by Hadley Wickham and Garrett Grolemund at http://r4ds.had.co.nz/

- Foundations of Statistics with R by Darrin Speegle: https://bookdown.org/speegled/foundations-ofstatistics/

- Repository of Bookdowns on a variety of topics related to data analysis, R, and statistics:
https://bookdown.org/home/tags/statistics/


### Territorial acknowledgement

UNBC is located on and serves diverse traditional territories that are home to numerous First Nations and other Aboriginal groups. The Prince George campus is situated on the traditional territory of the Lheidli T’enneh, part of the Dakelh (Carrier) First Nations. For further information about Territories on which UNBC campuses are located, please see: https://www2.unbc.ca/indigenous-resource-dati/traditional-territory-acknowledgement


### Academic accommodations

Students with diverse learning styles and needs are welcome in this course. If you have a disability or health condition that may require accommodations, please feel free to approach me and/or the UNBC Access Resource Centre (ARC) as soon as possible. The Access Resource Centre (ARC) provides service to students with documented disabilities or health conditions, ranging from permanent to temporary, including but not limited to chronic health issues, hearing and visual impairments, learning disabilities and attention deficit/hyperactivity disorders, mental health and neurological disabilities, and mobility and other physical disabilities. ARC staff are available by appointment to assess specific needs, provide referrals, and arrange appropriate academic accommodations to assist students in achieving their academic goals. Students who may have a need for academic accommodation are encouraged to contact ARC by email at arc@unbc.ca, by phone at 250-960-5682 (toll free 1-888-960-5682), or in person at 5- 157. For more information, please visit the Access Resource Centre website.

### Academic dishonesty

University regulations strictly forbid academic dishonesty of any type, including plagiarism, cheating during exams, or misrepresenting the nature of your involvement in any assigned work. Students involved in such acts may receive an automatic F in the course. Please consult the “Regulations and Policies” section of the UNBC Undergraduate Calendar if you are unfamiliar with definitions of any of the Academic Offences. Note that ignorance of these policies will not be accepted as a defense. Students guilty of such activities may receive a zero on the particular assignment/exam/lab in question. If deemed sufficiently severe by the instructor, the offense could lead to sanctions that include reprimand, reduction of grade, or suspension. All matters of Academic Misconduct at UNBC require instructors to submit a formal report to the Chair and Office of Registrar, which will be placed in the student’s academic file.
